# Create a crab detector

The folder `create-a-crab-detector` contains sripts and datasets which makes its possible to go from raw annotations to a trained YOLO-model. The first script
(`1-raw-to-yolo.ipnb`) takes raw CVAT annotations and turns them into a dataset ready for training an YOLO object detection model: RLE mask are converted into bounding boxes, and the data is split into five stratified folds for cross-validation. 

If you're applying this dataset to other than the shipped example, keep in mind (but not limited to); file paths: `CONFIG` needs to point at your own annotation JSON and image folder, and stratified sampling for the k-folds, this assumes your images can be grouped by BRUV station and labeled crab-present/empty; if that doesn't apply to your data, this step may need adjusting or turning off (see Chunk 9).

By the end of the script, each image should have a matching `.txt` file listing its bounding box annotations (or an empty file, if there's no crab in it), organized into five train/validation folds ready for object detection training.

## From raw annotations to a YOLO model
### 1. Setup and loading data (Chunk 0 -- 3) {-}

These chunks handle setup: installing any missing Python packages, loading the required libraries, and locating your dataset. They should run without errors, if Chunk 3 can't find your annotation file or image folder, double check the paths.

### 2. Converting masks to bounding boxes (Chunk 4) {-}

RLE masks are precise, but training directly on them takes more compute time and is really built for image segmentation, not simple object detection. Since the goal here is just to find and count crabs (not trace their exact outline), converting to bounding boxes is the better trade-off as they're faster to train on, and sufficient for detection. RLE masks are faster to draw as the annotation program utlized for the example datasets, Computer Vision Annotation Tool (CVAT;Sekachev,2020) applies Segment Anything Model 2 from META [@ravi_sam_2024] which makes the annotation process much faster than drawing traditional bounding boxes.

This chunk converts each RLE mask into a tightly-fitted bounding box, and normalizes the box coordinates to the 0–-1 range YOLO expects. Two filters are applied at the same time: annotations for any species other than decapod: crab are discarded, and overlapping duplicate boxes (drawn twice for the same crab) are reduced to one.

### 3. Linking images to annotations (Chunk 5) {-}

Links each annotation to its actual image file, matching by filename, again make sure image filenames match the identifiers in the annotation JSON exactly, any mismatches get reported as "unmatched," usually caused by a typo, a renamed file, or an image missing from the folder.

Note: frames with no crab in them (negative examples) are deliberately kept in the dataset rather than discarded, training on empty frames as well as crab-present ones has been shown to improve performance [@sorrenti_weakly_2025]. These get an empty `.txt` label file later, once the data is exported (Chunk 9).

### 4. Visual quality checks (Chunk 6 -- 7) {-}

These chunks are a visual quality check, not a processing step, and nothing here changes the data. Chunk 6 displays the first annotated image with its converted bounding box overlaid: every crab should have a red box tightly wrapping it, with no gaps or extra boxes.

Chunk 7 places the original RLE mask and the converted bounding box for the same crab side by side, to visually confirm the conversion in Chunk 4 didn't distort or misplace anything before relying on it for training.

### 5. Location and station helpers (Chunk 8) {-}

Filenames encode which location and BRUV station each image came from, but that information isn't stored as a separate field anywhere, it has to be pulled out of the filename itself. This chunk does that: `extract_location` reads out the site name (e.g. "Lovund"), and `extract_bruv_id` reads out the specific station ID (e.g. "LO3"). It also prints a quick summary: how many images come from each location, and how many of those are crab-present versus empty, worth a glance to check whether the dataset is skewed toward a few sites, since that affects how balanced the folds can be.

If applying another dataset with a different naming logic, these two functions will need to be adjusted to match the naming pattern (see Section \@ref(data).

### 6. Splitting into stratified, grouped folds (Chunk 9) {-}

This is the core of the export step: it splits the linked dataset into five folds for cross-validation, applying two rules at once.

Grouped: every image from the same BRUV station stays entirely in either training or validation within a fold, never split between the two. If the same station showed up in both, the model could effectively be tested on footage from a site it had already trained on, making it look like it's performing better than it actually would on a genuinely new site.

Stratified: within that constraint, each fold's train/validation split keeps a similar proportion of crab-present vs. empty frames as the full dataset, so no fold accidentally ends up training almost entirely on empty frames.

The function has a `stratify` option: `True` (the default) gives the behavior above; `False` skips the crab/empty balancing and keeps only the station grouping. Turning it off makes sense if your dataset is already fairly balanced, or if the crab/empty ratio varying between folds isn't a concern for you, station grouping still applies either way, so there's no risk of leakage regardless of this setting.

For each fold, this chunk writes the images and label files into `images/train`, `images/val`, `labels/train`, `labels/val`, along with a `dataset.yaml`.

YAML is a plain-text format for storing settings, written as simple, human-readable lines rather than code. YOLO's training tool reads `dataset.yaml` to know where to find the images and labels and what classes to expect. Each fold's file specifies: `path` (that fold's root folder), `train` and `val` (the image subfolders), `nc` (number of classes - 1, since this project only detects crabs), and `names` (mapping each class number to its name, e.g. `0: crab`).


### 7. Running the export (Chunk 10) {-}

This chunk runs the function from Chunk 9 on your dataset. By default it creates five folds under `data/kfold/`, each structured as:
```
data/kfold/
├── fold1/
│   ├── images/train/
│   ├── images/val/
│   ├── labels/train/
│   ├── labels/val/
│   └── dataset.yaml
├── fold2/ ... fold5/          # same structure as fold1
├── master_dataset.yaml
└── fold_summary.csv
```
along with a `master_dataset.yaml`, a small reference file at the top level listing the class setup (`nc` and `names`, same as each fold's file) and a comment reminding you how to point training at a specific fold (e.g. `yolo train data=fold1/dataset.yaml ...`), and a `fold_summary.csv` with the actual per-fold stats mentioned above.

Once this runs cleanly, no station-overlap warnings, and a crab/empty ratio that looks reasonable in the fold summary, the data is ready for training in `train.ipynb`.

A caveat is that with this limited dataset of only ten stations, we are not expecting the folds to be perfectly balanced, and it is expected that the folds that look more unbalanced will perform worse than the others. Now the dataset is ready for training!

```{r setup, include=FALSE}
library(kableExtra)
library(dplyr)
```
## Training the crab detector

This script (`train.ipynb`) takes the five-fold dataset produced by `1-raw_to_yolo.ipynb` and trains a YOLOv8 object detector on it, one model per fold, so performance can be evaluated across several independent train/validation splits rather than trusting a single one.

### 1. Setup (Chunk 1) {-}

This chunk locates the repository (the same `find_repo_root()` approach used in `1-raw_to_yolo.ipynb`, so paths work the same whether its running on a laptop or a server), sets up where training output gets written, and checks that a GPU is available. Training a model like this on a CPU is impractically slow, so this check is worth paying attention to: if `CUDA available: False` prints here, stop and sort out the GPU setup before continuing rather than letting a full training run crawl for days.

### 2. Checking the folds (Chunk 2) {-}

Before spending hours training, this chunk verifies every fold from `1-raw_to_yolo.ipynb` is actually complete: that each fold has a readable `dataset.yaml`, and that the number of images and label files line up in both the train and validation sets. This is a safety gate, not a formality, training on a broken or incomplete fold wastes real compute time, so it's worth letting this fail loudly here rather than partway through a long run. Chunk 5 (the real training) refuses to start if this check doesn't pass.

### 3. Hyperparameters (Chunk 3) {-}

This chunk sets out every training parameter in one place, so the whole run is reproducible from a single configuration. Rather than explaining every individual number, it's easier to think of them in three groups:

**Model and image handling.** `yolov8s.pt` is the pretrained starting point the model builds on rather than learning from scratch, you can freely try other sizes of YOLOv8 or other versions by changing them out here. Images are resized to 960 pixels and processed in batches of 16, larger images and batches use more GPU memory, but a larger image size in particular can help the model pick out small objects like crabs against a wide frame.

**How the model learns.** The optimizer, learning rate, and warmup settings control how aggressively the model adjusts itself while training, and how that pace changes over the course of the run (a "cosine" schedule, `cos_lr`, means the learning rate eases off smoothly toward the end rather than dropping abruptly). These values were carried over from what worked in the original thesis training runs, rather than re-tuned from scratch here.

**Augmentation.** During training, images are randomly rotated, shifted, scaled, and mirrored slightly before being shown to the model. The intent isn't to create unrealistic images, it's to stop the model from memorizing the exact footage it was trained on, so it generalizes better to new footage. A couple of these are deliberately constrained for this dataset: `flipud=0` means images are never flipped upside-down (a crab is never upside-down in real footage, so training on flipped versions would teach the model the wrong thing), and `single_cls=True` reflects that there's only one class to detect here (crab), so the model doesn't spend effort distinguishing between species that don't apply.

Two epoch counts are set here: `test_epochs` (5) for the quick sanity check in Chunk 4, and `epochs` (100) for the real run in Chunk 5, 100 was chosen because performance had already converged by that point in the original thesis training.

### 4. Test run (Chunk 4) {-}

This runs a handful of epochs on one fold using the exact same settings as the full run, purely to confirm the pipeline works end-to-end, paths resolve, the GPU is used correctly, and training completes without errors before committing to a run that takes hours. The actual metrics from this run aren't meaningful and shouldn't be read as a preview of model quality: five epochs isn't enough for the model to properly learn, and results can look noisy or even get temporarily worse epoch to epoch. That's expected here, not a warning sign.

### 5. Full training (Chunk 5) {-}

This trains one full model per fold, using the hyperparameters from Chunk 3. Each fold's best-performing checkpoint is saved automatically, along with a checkpoint every 25 epochs during training in case a run needs to be resumed. There's a short pause between folds to let the GPU cool down. The current server utilized is NVIDIA RTX A6000, which used around 3-4 hours for the entire training run, whilst training the full dataset used in the thesis took around 16--18 hours.  

### 6. Collecting and interpreting results (Chunks 6 -- 8) {-}

#### Collection {-}
Training five separate models means five separate sets of results, so these chunks pull them together: loading each fold's training log, plotting the key metrics (mAP, precision, recall) side by side across folds, and then computing a summary, both the final numbers from training and a fresh, explicit validation pass on each fold's saved model, as a double-check against the training log alone.

The reason to look across all folds rather than picking one is the same logic behind K-fold cross-validation in general: any single fold could happen to get an easy or a hard validation split by chance, so the story that matters is the overall pattern, not one number in isolation.

#### Interpretation {-}
One limitation worth being upfront about: this dataset has only 10 BRUV stations, and because stations are kept fully separate between training and validation (to avoid the model being tested on footage it effectively already trained on -- see the K-fold export step in `raw_to_yolo.ipynb`), each fold's validation set ends up being just 1-3 stations. That's a small, uneven sample, and fold-to-fold variation in these results should be read as a consequence of the dataset's size, not a flaw in the training method. One fold in particular had very few crab-positive images in its validation set purely by chance, and its numbers should be treated as unreliable rather than averaged in on equal footing with the others.

To evaluate object detection model performance, standard object detection metrics were used: true positives, false positives, false negatives, mean average precision (mAP50), precision, recall, and F1 (Table \@ref(tab:cnnMetricsdef)). After training, each model was validated on the validation dataset. The model predicts the class and location of objects in each frame with bounding boxes. To assign these metrics, predictions were ranked by confidence and matched to the manually annotated bounding boxes from the validation dataset, with each annotation matched to at most one prediction. A predicted box was counted as a true positive (TP) if it overlapped an annotation of the same class sufficiently, assessed using intersection-over-union (IoU); unmatched predictions were counted as false positives (FP) and unmatched annotations as false negatives (FN) (a confusion matrix showing the distirbution of all these can be found in `test_runs/models/fold1_yolov8s_100epochs`. IoU is calculated as the area of overlap between the predicted and annotated box divided by the area of their union, giving a score between 0 and 1 [@biscarini_deep_2026]. An IoU threshold of 0.5 was applied, meaning a prediction had to overlap at least 50% with the annotated box to be counted as correct, following standard object detection practice.

```{r cnnMetricsdef, echo = FALSE}
cnn_metrics <- data.frame(
  Metric = c("True positive (TP)", "False positive (FP)", "False negative (FN)",
             "Precision (0--1)", "Recall (0--1)", "F1 (0--1)",
             "Mean average precision (mAP50)"),
  Description = c(
    "Correct detections: predicted bounding box matches an annotated box.",
    "Model predicts a bounding box where there is no object.",
    "An object is present but the model produces no matching prediction.",
    "Ratio of correct positive predictions to all positive predictions, measures detection accuracy.",
    "Ratio of correct positive predictions to all actual positives, measures the model's ability to find all objects.",
    "Harmonic mean of precision and recall, balances avoiding false detections with capturing all objects.",
    "Average of precision scores across recall levels, summarized across all classes ($C$)."
  ),
Formula = c("", "", "",
  "TP / (TP + FP)",
  "TP / (TP + FN)",
  "2 * TP / (2 * TP + FP + FN)",
  "(1/C) * sum(AP\\c) for c = 1...C")
)
cnn_metrics %>%
  kbl(
    col.names = c("Metric", "Description", "Formula"),
    booktabs  = TRUE,
    escape    = FALSE,
    caption   = "Standard performance metrics for assessing object detection models. A detection is counted as correct (TP) if the predicted bounding box overlaps with an annotated box by at least 50\\% (IoU $\\geq$ 0.5). Metrics range from 0 to 1, where 1 indicates perfect performance.",
    align     = c("l", "l", "c")
  ) 
```

Aside for interpreting the performance metrics, its worth checking out the visual validation confirmations. In the folder `test_runs/models/fold1_yolov8s_100epochs`, or any other fold. Compare the `val_batch_labels.jpg` with `val_batch_pred.jpg`, this compares the actual annotated images and the predicted annotations from the model to see the confidence scores and what the model catches/misses. If the behavior seems wrong, there is a plethora of tweaks possible to improve model behavior that I wont touch on here but is easiliy available online or with the help of LLMs.